---
title: "GPS Data Visualization Analysis"
author: "GPS Analysis Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(ggplot2)
library(plotly)
library(DT)
library(knitr)
library(kableExtra)

# Settings
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Colors
theme_set(theme_minimal() + 
          theme(plot.title = element_text(size = 14, face = "bold"),
                axis.title = element_text(size = 12),
                legend.title = element_text(size = 11)))
```

# Data Loading

```{r load-data}
# Load master CSV from GitHub raw URL
gps_data <- read_csv("https://raw.githubusercontent.com/tomaszbielNCI/kml_procesor/master/data/output/gps_master.csv")

# Convert dates - handle different time formats
gps_data <- gps_data %>%
  mutate(
    # Handle time from GPX extensions (HH:MM:SS format)
    point_clock_time = ifelse(point_clock != "no_time" & !is.na(point_clock), 
                             paste(route_date, point_clock), NA),
    
    # Convert standard time
    time_standard = ifelse(time != "no_time" & !grepl("^[0-9]{2}:[0-9]{2}:[0-9]{2}$", time), 
                         time, NA),
    
    # Create one time column
    time_clean = coalesce(time_standard, point_clock_time),
    
    # Convert to datetime
    time = ymd_hms(time_clean),
    
    date = date(time),
    hour = hour(time),
    month = month(time, label = TRUE),
    file_type = factor(file_type)
  )

# Basic statistics
cat("Number of GPS points:", nrow(gps_data), "\n")
cat("Number of unique routes:", n_distinct(gps_data$track_name), "\n")
cat("Date range:", min(gps_data$date, na.rm = TRUE), "-", max(gps_data$date, na.rm = TRUE), "\n")
cat("Points with time:", sum(!is.na(gps_data$time)), "\n")
cat("Points with clock:", sum(!is.na(gps_data$point_clock)), "\n")
```

---

## 1. Altitude Histogram

```{r plot1}
ggplot(gps_data, aes(x = altitude)) +
  geom_histogram(bins = 50, fill = "skyblue", alpha = 0.7, color = "black") +
  labs(title = "Distribution of GPS Point Altitudes",
       x = "Altitude (m above sea level)", y = "Number of Points") +
  geom_vline(aes(xintercept = mean(altitude, na.rm = TRUE)), 
             color = "red", linetype = "dashed", size = 1)
```

---

## 2. Altitude Boxplot by Routes

```{r plot2}
top_routes <- gps_data %>% 
  count(track_name, sort = TRUE) %>% 
  slice_head(n = 10) %>% 
  pull(track_name)

gps_data %>% 
  filter(track_name %in% top_routes) %>%
  ggplot(aes(x = reorder(track_name, altitude, median), y = altitude)) +
  geom_boxplot(fill = "lightgreen", alpha = 0.7) +
  coord_flip() +
  labs(title = "Altitude Distribution for Top 10 Routes",
       x = "Route", y = "Altitude (m above sea level)")
```

---

## 3. Latitude Distribution

```{r plot3}
ggplot(gps_data, aes(x = latitude)) +
  geom_density(fill = "orange", alpha = 0.6) +
  labs(title = "Density Distribution of Latitude",
       x = "Latitude", y = "Density")
```

---

## 4. Longitude Distribution

```{r plot4}
ggplot(gps_data, aes(x = longitude)) +
  geom_density(fill = "purple", alpha = 0.6) +
  labs(title = "Density Distribution of Longitude",
       x = "Longitude", y = "Density")
```

---

## 5. Altitude vs Latitude

```{r plot5}
ggplot(gps_data, aes(x = latitude, y = altitude)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  labs(title = "Altitude vs Latitude Relationship",
       x = "Latitude", y = "Altitude (m above sea level)")
```

---

## 6. Speed Distribution

```{r plot6}
# Calculate speed from position changes for all data
speed_data <- gps_data %>%
  arrange(track_name, time) %>%
  group_by(track_name) %>%
  mutate(
    lat_diff = latitude - lag(latitude),
    lon_diff = longitude - lag(longitude),
    time_diff = as.numeric(difftime(time, lag(time), units = "secs")),
    distance_km = sqrt(lat_diff^2 + lon_diff^2) * 111,
    speed_kmh = ifelse(time_diff > 0, (distance_km / time_diff) * 3600, NA)
  ) %>%
  filter(!is.na(speed_kmh) & speed_kmh < 150) %>%
  ungroup()

ggplot(speed_data, aes(x = speed_kmh)) +
  geom_histogram(bins = 50, fill = "darkgreen", alpha = 0.7) +
  labs(title = "Speed Distribution Across All Routes",
       x = "Speed (km/h)", y = "Number of Points") +
  geom_vline(aes(xintercept = mean(speed_kmh, na.rm = TRUE)), 
             color = "red", linetype = "dashed", size = 1)
```

---

## 7. Time Distribution of Points (Hours)

```{r plot7}
gps_data %>%
  filter(!is.na(hour)) %>%
  ggplot(aes(x = hour)) +
  geom_histogram(bins = 24, fill = "gold", alpha = 0.7, color = "black") +
  labs(title = "Distribution of GPS Points by Hour of Day",
       x = "Hour", y = "Number of Points") +
  scale_x_continuous(breaks = seq(0, 24, 2))
```

---

## 8. Points per Route

```{r plot8}
route_counts <- gps_data %>% 
  count(track_name, sort = TRUE) %>% 
  slice_head(n = 15)

ggplot(route_counts, aes(x = reorder(track_name, n), y = n)) +
  geom_col(fill = "coral", alpha = 0.8) +
  coord_flip() +
  labs(title = "GPS Points for 15 Longest Routes",
       x = "Route", y = "Number of Points")
```

---

## 9. Average Altitude per Route

```{r plot9}
route_elevation <- gps_data %>% 
  group_by(track_name) %>% 
  summarise(mean_elevation = mean(altitude, na.rm = TRUE),
            .groups = 'drop') %>% 
  arrange(desc(mean_elevation)) %>% 
  slice_head(n = 15)

ggplot(route_elevation, aes(x = reorder(track_name, mean_elevation), y = mean_elevation)) +
  geom_col(fill = "brown", alpha = 0.8) +
  coord_flip() +
  labs(title = "Average Altitude for 15 Highest Routes",
       x = "Route", y = "Average Altitude (m above sea level)")
```

---

## 10. GPS Points Scatter Map

```{r plot10}
ggplot(gps_data, aes(x = longitude, y = latitude)) +
  geom_point(alpha = 0.4, color = "navy", size = 0.5) +
  labs(title = "GPS Points Scatter Map",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

---

## 11. Altitude vs Time (for time-series data)

```{r plot11}
gps_data %>%
  filter(!is.na(time)) %>%
  slice_sample(n = 5000) %>%  # Sampling for performance
  ggplot(aes(x = time, y = altitude, color = track_name)) +
  geom_line(alpha = 0.7) +
  labs(title = "Altitude Changes Over Time",
       x = "Time", y = "Altitude (m above sea level)") +
  theme(legend.position = "none")
```

---

## 12. Altitude Statistics by File Types

```{r plot12}
ggplot(gps_data, aes(x = file_type, y = altitude, fill = file_type)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Altitude Distribution by File Types",
       x = "File Type", y = "Altitude (m above sea level)") +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "none")
```

---

## 13. 2D Point Density (Contour)

```{r plot13}
ggplot(gps_data, aes(x = longitude, y = latitude)) +
  geom_density_2d_filled(alpha = 0.8) +
  labs(title = "GPS Points Density (2D Contour)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

---

## 14. Geographic Range of Each Route

```{r plot14}
route_bounds <- gps_data %>%
  group_by(track_name) %>%
  summarise(
    lat_range = max(latitude, na.rm = TRUE) - min(latitude, na.rm = TRUE),
    lon_range = max(longitude, na.rm = TRUE) - min(longitude, na.rm = TRUE),
    total_range = lat_range + lon_range,
    .groups = 'drop'
  ) %>%
  arrange(desc(total_range)) %>%
  slice_head(n = 15)

ggplot(route_bounds, aes(x = reorder(track_name, total_range), y = total_range)) +
  geom_col(fill = "darkblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Geographic Range of 15 Largest Routes",
       x = "Route", y = "Geographic Range (degrees)")
```

---

## 15. Correlation Between Coordinates

```{r plot15}
gps_data %>%
  sample_n(10000) %>%  # Sampling
  ggplot(aes(x = latitude, y = longitude)) +
  geom_point(alpha = 0.3, color = "red") +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  labs(title = "Correlation Between Latitude and Longitude",
       x = "Latitude", y = "Longitude")
```

---

## 16. Altitude Distribution with Quartiles

```{r plot16}
ggplot(gps_data, aes(x = altitude)) +
  geom_histogram(bins = 50, fill = "lightblue", alpha = 0.7) +
  geom_vline(aes(xintercept = quantile(altitude, 0.25, na.rm = TRUE)), 
             color = "orange", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = quantile(altitude, 0.5, na.rm = TRUE)), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = quantile(altitude, 0.75, na.rm = TRUE)), 
             color = "purple", linetype = "dashed", size = 1) +
  labs(title = "Altitude Distribution with Quartile Markers",
       x = "Altitude (m above sea level)", y = "Number of Points") +
  annotate("text", x = Inf, y = Inf, 
           label = "Q1 | Q2 | Q3", 
           hjust = 1.1, vjust = 1.1,
           color = c("orange", "red", "purple"))
```

---

## 17. Route Statistics Summary Table

```{r plot17}
route_stats <- gps_data %>%
  group_by(track_name) %>%
  summarise(
    points = n(),
    avg_altitude = round(mean(altitude, na.rm = TRUE), 1),
    min_altitude = round(min(altitude, na.rm = TRUE), 1),
    max_altitude = round(max(altitude, na.rm = TRUE), 1),
    altitude_range = round(max_altitude - min_altitude, 1),
    calories = round(mean(route_total_calories, na.rm = TRUE), 0),
    distance_km = round(mean(route_distance_km, na.rm = TRUE), 2),
    file_type = first(file_type),
    .groups = 'drop'
  ) %>%
  arrange(desc(points)) %>%
  slice_head(n = 20)

DT::datatable(route_stats, 
              options = list(pageLength = 10, scrollX = TRUE),
              caption = "Statistics for 20 Largest Routes")
```

---

## 18. Time Analysis from GPX Extensions

```{r plot18}
# Data with clock time
time_analysis <- gps_data %>%
  filter(!is.na(point_clock)) %>%
  mutate(
    time_seconds = as.numeric(point_seconds),
    time_minutes = time_seconds / 60
  )

ggplot(time_analysis, aes(x = time_minutes)) +
  geom_histogram(bins = 50, fill = "lightgreen", alpha = 0.7) +
  labs(title = "Distribution of Route Time (minutes from start)",
       x = "Time (minutes)", y = "Number of Points") +
  geom_vline(aes(xintercept = mean(time_minutes, na.rm = TRUE)), 
             color = "red", linetype = "dashed", size = 1)
```

---

## 19. Calories vs Distance

```{r plot19}
# Force convert columns to numeric and remove NAs
route_total_calories <- as.numeric(gps_data$route_total_calories)
route_distance_km <- as.numeric(gps_data$route_distance_km)

# Create data frame for plotting
calories_distance_data <- data.frame(
  calories = route_total_calories[!is.na(route_total_calories) & !is.na(route_distance_km) & route_total_calories > 0],
  distance = route_distance_km[!is.na(route_total_calories) & !is.na(route_distance_km) & route_distance_km > 0]
)

# Check data
cat("Valid data points:", nrow(calories_distance_data), "\n")

if(nrow(calories_distance_data) > 0) {
  ggplot(calories_distance_data, aes(x = distance, y = calories)) +
    geom_point(alpha = 0.6, color = "darkred") +
    geom_smooth(method = "lm", color = "blue", se = TRUE) +
    labs(title = "Calories vs Distance Relationship",
         x = "Distance (km)", y = "Calories") +
    annotate("text", x = Inf, y = -Inf, 
             label = paste("Correlation:", round(cor(distance, calories, use = "complete.obs"), 2)),
             hjust = 1.1, vjust = -0.5)
} else {
  cat("No valid data for plotting")
}
```

---

## 20. Valleymount to Blessington - Time vs Speed Analysis (Anscombe's Quartet)

```{r plot20}
# Filter Valleymount to Blessington route
valleymount_route <- gps_data %>%
  filter(track_name == "valleymount to blessington  12/10/2025 09:22") %>%
  filter(!is.na(point_seconds)) %>%
  mutate(
    time_minutes = as.numeric(point_seconds) / 60,
    # Calculate speed from position changes
    lat_diff = latitude - lag(latitude),
    lon_diff = longitude - lag(longitude),
    time_diff = as.numeric(point_seconds) - lag(as.numeric(point_seconds)),
    distance_km = sqrt(lat_diff^2 + lon_diff^2) * 111,  # Approximate km per degree
    speed_kmh = ifelse(time_diff > 0, (distance_km / time_diff) * 3600, 0),
    speed_kmh = ifelse(speed_kmh > 150, NA, speed_kmh)  # Remove unrealistic speeds
  ) %>%
  filter(!is.na(speed_kmh))

# Time vs Speed scatter plot with regression line
ggplot(valleymount_route, aes(x = time_minutes, y = speed_kmh)) +
  geom_point(alpha = 0.6, color = "orange") +
  geom_smooth(method = "lm", color = "red", se = TRUE, linewidth = 1) +
  labs(title = "Valleymount to Blessington - Time vs Speed Analysis",
       subtitle = "Showing transition from walking to car journey",
       x = "Time (minutes)", y = "Speed (km/h)") +
  theme_minimal() +
  annotate("text", x = Inf, y = Inf, 
           label = paste("R² =", round(summary(lm(speed_kmh ~ time_minutes, data = valleymount_route))$r.squared, 3)),
           hjust = 1.1, vjust = 1.1, color = "red")
```

---

## 21. Valleymount to Blessington - Speed Phases Analysis

```{r plot21}
# Identify walking vs car phases (75% walking, 25% car)
walking_phase <- valleymount_route %>% head(round(nrow(valleymount_route) * 0.75))
car_phase <- valleymount_route %>% tail(round(nrow(valleymount_route) * 0.25))

# Combine with phase labels
speed_analysis <- bind_rows(
  walking_phase %>% mutate(phase = "Walking"),
  car_phase %>% mutate(phase = "Car")
)

# Box plot comparison
ggplot(speed_analysis, aes(x = phase, y = speed_kmh, fill = phase)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Speed Comparison: Walking vs Car Phases",
       subtitle = "Valleymount to Blessington Route",
       x = "Journey Phase", y = "Speed (km/h)") +
  scale_fill_manual(values = c("Walking" = "lightblue", "Car" = "lightcoral")) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## 22. Valleymount to Blessington - Speed Timeline with Transition

```{r plot22}
# Create speed timeline with transition point
transition_time <- max(walking_phase$time_minutes)

ggplot(valleymount_route, aes(x = time_minutes, y = speed_kmh)) +
  geom_line(alpha = 0.7, color = "steelblue") +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_vline(xintercept = transition_time, color = "red", linetype = "dashed", linewidth = 1.5) +
  annotate("text", x = transition_time, y = max(valleymount_route$speed_kmh, na.rm = TRUE) * 0.9,
           label = "Transition to Car", angle = 90, vjust = -0.5, color = "red") +
  labs(title = "Speed Timeline - Valleymount to Blessington Route",
       subtitle = "Clear transition point where walking changes to car journey",
       x = "Time (minutes)", y = "Speed (km/h)") +
  theme_minimal()
```

---

## 23. Violin Plot of Speed by Route Type

```{r plot23}
# Calculate speed for all data first
gps_data_with_speed <- gps_data %>%
  arrange(track_name, time) %>%
  group_by(track_name) %>%
  mutate(
    lat_diff = latitude - lag(latitude),
    lon_diff = longitude - lag(longitude),
    time_diff = as.numeric(difftime(time, lag(time), units = "secs")),
    distance_km = sqrt(lat_diff^2 + lon_diff^2) * 111,
    speed_kmh = ifelse(time_diff > 0, (distance_km / time_diff) * 3600, NA)
  ) %>%
  filter(!is.na(speed_kmh) & speed_kmh < 150) %>%
  ungroup()

# Categorize routes by type (hiking vs transport)
route_categories <- gps_data_with_speed %>%
  mutate(
    route_type = case_when(
      grepl("walk|hiking|mountain", track_name, ignore.case = TRUE) ~ "Hiking",
      grepl("car|drive|blessington", track_name, ignore.case = TRUE) ~ "Transport",
      TRUE ~ "Other"
    )
  )

ggplot(route_categories, aes(x = route_type, y = speed_kmh, fill = route_type)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, alpha = 0.8, color = "black") +
  labs(title = "Speed Distribution by Route Type",
       x = "Route Type", y = "Speed (km/h)") +
  scale_fill_manual(values = c("Hiking" = "lightgreen", "Transport" = "lightcoral", "Other" = "lightblue")) +
  theme_minimal()
```

---

## 24. Bubble Plot - Distance vs Points vs Calories

```{r plot24}
# Create bubble plot data with proper calculations
bubble_data <- gps_data %>%
  group_by(track_name) %>%
  summarise(
    distance = mean(route_distance_km, na.rm = TRUE),
    points = n(),
    calories = mean(route_total_calories, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  filter(!is.na(distance) & !is.na(calories))

ggplot(bubble_data, aes(x = distance, y = points, size = calories, color = calories)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(3, 15), name = "Calories") +
  scale_color_gradient(low = "lightblue", high = "darkred", name = "Calories") +
  labs(title = "Bubble Plot: Distance vs Points vs Calories",
       x = "Distance (km)", y = "Number of Points") +
  theme_minimal()
```

---

## 25. Stacked Bar Chart - File Types by Route

```{r plot25}
# Create stacked data
stacked_data <- gps_data %>%
  count(track_name, file_type) %>%
  group_by(file_type) %>%
  slice_max(n = 10) %>%
  ungroup() %>%
  mutate(track_name = factor(track_name, levels = unique(track_name)))

ggplot(stacked_data, aes(x = track_name, y = n, fill = file_type)) +
  geom_col(position = "fill") +
  coord_flip() +
  labs(title = "Stacked Bar Chart: File Types by Route (Top 10)",
       x = "Route", y = "Number of Points", fill = "File Type") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## 26. Heat Map - Activity by Hour and Day

```{r plot26}
# Create heat map data
heat_data <- gps_data %>%
  filter(!is.na(hour) & !is.na(date)) %>%
  mutate(day = wday(date, label = TRUE, week_start = 1)) %>%
  count(day, hour) %>%
  complete(day, hour, fill = 0)

ggplot(heat_data, aes(x = hour, y = day, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred", name = "Points") +
  labs(title = "Heat Map: GPS Activity by Hour and Day",
       x = "Hour of Day", y = "Day of Week") +
  theme_minimal()
```

---

## 27. Trellis Plot - Altitude by Route

```{r plot27}
# Create trellis (facet) plot with proper data
trellis_data <- gps_data %>%
  filter(!is.na(altitude)) %>%
  arrange(track_name, time) %>%
  group_by(track_name) %>%
  mutate(
    point_index = row_number(),
    time_minutes = as.numeric(difftime(time, min(time, na.rm = TRUE), units = "mins"))
  ) %>%
  ungroup() %>%
  filter(track_name %in% top_routes) %>%
  sample_n(1000)  # Sample for clarity

ggplot(trellis_data, aes(x = time_minutes, y = altitude)) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  facet_wrap(~track_name, scales = "free_y") +
  labs(title = "Trellis Plot: Altitude Profiles by Route",
       x = "Time (minutes)", y = "Altitude (m)") +
  theme_minimal()
```

---

## 28. Strip Chart with Means

```{r plot28}
# Strip chart with means
strip_data <- gps_data %>%
  filter(!is.na(route_total_calories)) %>%
  group_by(track_name) %>%
  summarise(
    calories = mean(route_total_calories, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(calories)) %>%
  head(15)

ggplot(strip_data, aes(x = track_name, y = calories)) +
  geom_jitter(width = 0.2, alpha = 0.6, color = "darkblue") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, color = "red") +
  coord_flip() +
  labs(title = "Strip Chart with Means: Average Calories by Route",
       x = "Route", y = "Average Calories") +
  theme_minimal()
```

---

# Summary

```{r summary}
cat("=== GPS DATA SUMMARY ===\n\n")
cat("Total number of points:", nrow(gps_data), "\n")
cat("Number of unique routes:", n_distinct(gps_data$track_name), "\n")
cat("Number of GPX files:", sum(gps_data$file_type == "gpx"), "\n")
cat("Number of KML files:", sum(gps_data$file_type == "kml"), "\n")
cat("Points with time:", sum(!is.na(gps_data$time)), "\n")
cat("Points with clock:", sum(!is.na(gps_data$point_clock)), "\n")
cat("Altitude range:", round(min(gps_data$altitude, na.rm = TRUE), 1), "-", 
    round(max(gps_data$altitude, na.rm = TRUE), 1), "m above sea level\n")
cat("Average altitude:", round(mean(gps_data$altitude, na.rm = TRUE), 1), "m above sea level\n")
cat("Latitude range:", round(min(gps_data$latitude, na.rm = TRUE), 4), "-", 
    round(max(gps_data$latitude, na.rm = TRUE), 4), "\n")
cat("Longitude range:", round(min(gps_data$longitude, na.rm = TRUE), 4), "-", 
    round(max(gps_data$longitude, na.rm = TRUE), 4), "\n")

if(sum(!is.na(gps_data$route_total_calories)) > 0) {
  cat("Calories range:", round(min(gps_data$route_total_calories, na.rm = TRUE)), "-", 
      round(max(gps_data$route_total_calories, na.rm = TRUE)), "kcal\n")
  cat("Average calories:", round(mean(gps_data$route_total_calories, na.rm = TRUE)), "kcal\n")
}

if(sum(!is.na(gps_data$route_distance_km)) > 0) {
  cat("Distance range:", round(min(gps_data$route_distance_km, na.rm = TRUE), 2), "-", 
      round(max(gps_data$route_distance_km, na.rm = TRUE), 2), "km\n")
  cat("Average distance:", round(mean(gps_data$route_distance_km, na.rm = TRUE), 2), "km\n")
}
```

---

*Generated on `r Sys.Date()` using R Markdown and ggplot2*
